Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts

### 2024-02-26 manuscript_elsevier/manuscript.Rmd
abstract

---
title: Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
# author:
#   - name: Damien Dupré
#     email: damien.dupre@dcu.ie
#     affiliation: Business School, Dublin City University
#     correspondingauthor: true
#   - name: Edgar Morgenroth
#     email: edgar.morgenroth@dcu.ie
#     affiliation: Business School, Dublin City University
# address:
#   - code: Business School, Dublin City University
#     organization: Business School, Dublin City University
#     city: Dublin
#     country: Ireland
abstract: |
  This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups. The study also evaluates the effects of school closures on COVID-19 cases using a Generalized Additive Model (GAM). The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied. However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases. Age groups 15 to 24 show a surge immediately after closure, followed by a decline. Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa. These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
keywords: 
  - COVID-19
  - School Closures
  - Intergenerational Transmission.
journal: "Public Health"
date: "`r Sys.Date()`"
classoption: preprint, 3p
linenumbers: false
numbersections: true
bibliography: bibliography.bib
# biblio-style: elsarticle-num
csl: https://www.zotero.org/styles/elsevier-with-titles
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
    keep_tex: false
    citation_package: default
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{caption}
---

@@

Recently, a body of literature has shown various negative consequences on child development and health due to school closures. School closures have been found to have negatively affected child mental health [@moulin2022longitudinal], nutrition/obesity [@sugimoto2023temporal], and education [@lerkkanen2023reading]. Some governments are now questioning their decisions to close schools [@de2021determines]. For example, the German health minister Karl Lauterbach, in an interview with one of Germany’s publicly funded television stations, admitted that “in retrospect it had been wrong to keep schools and childcare closed for so long” [@ard2023lauterbach].

Even if school closures had negative consequences for children, they might nevertheless have been effective at stopping the spread of COVID-19. However, analysis on the effectiveness of school closures on the spread of COVID-19 remains inconclusive. @alfano2022effects in a study of a panel of European countries found that school closures were associated with a reduced COVID-19 incidence. In contrast, @walsh2021school in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60% having identified no impact and pointing to the potential for the analysis to be affected by confounding factors and collinearity. The latter shortcomings suggest that it is hard to draw firm inferences, as even a positive result may not imply causality.

This paper investigates the dynamics of COVID-19 across age groups and how this was affected by school closures. This allows us to identify the effect of school closure on the evolution of COVID-19 cases in other age groups and particularly on older and more vulnerable age groups, including parents, grandparents, and relatives about which little is known.

We identify the relationship between the school closures and case numbers in different age groups using Generalized Additive Models (GAM), which are more flexible than conventional Vector Autoregressions, given that they allow for non-linear relationships between variables. Furthermore, compared to other methods such as Vector Autoregression, GAM can handle fixed and random smooth effects without making strict assumptions about linearity. Then, using the transfer entropy method, we evaluate the impact of case number changes between age groups in order to identify if school closure has secondary repercussions in the household. Transfer entropy can adapt to changes in the underlying system dynamics over time. It does not assume stationarity and can capture evolving causal relationships, making it suitable for studying dynamic systems and non-stationary processes such as COVID-19 contaminations.

## Influence of School Closure

@@

## Causal Relationship Between Age Groups

A second implicit belief regarding the effect of school closure on the spread of COVID-19 is that school not only influences the spread of the virus among children and teenagers but also has a knock-on effect on the spread of the virus in older age groups, also called Secondary Attack Rate (SAR). The contaminated children and teenagers would bring the virus back home and would pass the virus on to their parents and relatives. For example, research investigating the contamination in the household network not only revealed an exceptionally high rate of secondary contamination but also that this contamination happened when the schools were closed [@sorianoarandes2021household].

Despite being reported in several clinical and epidemiological studies [@siebach2021childhood; @zhendong2020clinical], multiple research have shown that the SAR from children to household members was lower than expected [@heavey2020evidence; @vanderhoek2020role; @kim2021role; @ludvigsson2020children]. However, the SAR of children and teenagers to the household member is likely to be age-dependent, with differences between infants, primary and secondary school children, and college students [@grasleguen2021reopening]. If a secondary transmission from children and teenagers to household member has a significant influence, then a temporal causality relationship between their evolution should be observed.

@@

# Estimation Methodology

## Generalized Additive Model

@@

## Transfer Entropy

@@

# Data

## COVID-19 cases per age group

For the analysis, data on COVID case counts by age group are required. The number of COVID-19 cases per age group has been taken from the COVerAGE-DB project [@riffe2021data]. The COVerAGE-DB project consists of 3 data files: an “Input” data file that collects the official COVID-19 cases from 117 countries, the “Output_5” data file, which is a projection of COVID-19 cases by a group of 5 years, and the “Output_10” data file, which is a projection of COVID-19 cases by a group of 10 years. For our analyses, we have used the “Output_5,” which deals with the heterogeneity of countries’ reporting formats by using spline approximations when the data for this age bracket is not available for a country. Therefore, it allows a precise analysis between age groups under 24, while age groups of 25 and above are reduced to 10-year brackets, and groups above 64 are concatenated all together.

## School closure

@@

# Discussion & Conclusion

@@

# Data Availability

All data extracted from the COVerAGE-DB project and from the UNESCO global education coalition that are used to produce these results are fully available here https://github.com/damien-dupre/covid_ts_causality. The repository also includes the R code that were used to produce these results.

>>>>

---
title: Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
# author:
#   - name: Damien Dupré
#     email: damien.dupre@dcu.ie
#     affiliation: Business School, Dublin City University
#     correspondingauthor: true
#   - name: Edgar Morgenroth
#     email: edgar.morgenroth@dcu.ie
#     affiliation: Business School, Dublin City University
# address:
#   - code: Business School, Dublin City University
#     organization: Business School, Dublin City University
#     city: Dublin
#     country: Ireland
abstract: |
  *Objectives:* This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups.

  *Study Design:* A longitudinal study was conducted to observe changes in COVID-19 cases during school closures.

  *Methods:* A Generalized Additive Model is performed to compare trends from twelve European countries across five age groups. Then, Transfer Entropy calculations are used to evaluate the impact of younger cohorts on older cohorts.

  *Results:* The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied. However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases. Age groups 15 to 24 show a surge immediately after closure, followed by a decline. Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa.

  *Conclusions:* These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
keywords: 
  - COVID-19
  - School Closures
  - Intergenerational Transmission.
journal: "Public Health"
date: "`r Sys.Date()`"
classoption: preprint, 3p
linenumbers: false
numbersections: false
bibliography: bibliography.bib
# biblio-style: elsarticle-num
csl: https://www.zotero.org/styles/elsevier-with-titles
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
    keep_tex: false
    citation_package: default
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{caption}
---

@@

Recently, a body of literature has shown various negative consequences on child development and health due to school closures. School closures have been found to have negatively affected child mental health [@moulin2022longitudinal], nutrition/obesity [@sugimoto2023temporal], and education [@lerkkanen2023reading]. Some governments are now questioning their decisions to close schools [@de2021determines]. For example, the German health minister Karl Lauterbach, in an interview with one of Germany’s publicly funded television stations, admitted that “in retrospect it had been wrong to keep schools and childcare closed for so long” [@ard2023lauterbach]. Even if school closures had negative consequences for children, they might nevertheless have been effective at stopping the spread of COVID-19. However, analysis on the effectiveness of school closures on the spread of COVID-19 remains inconclusive. @alfano2022effects in a study of a panel of European countries found that school closures were associated with a reduced COVID-19 incidence. In contrast, @walsh2021school in a review of 40 studies covering 150 countries concluded that the effectiveness of school closures was uncertain, with 60% having identified no impact and pointing to the potential for the analysis to be affected by confounding factors and collinearity. The latter shortcomings suggest that it is hard to draw firm inferences, as even a positive result may not imply causality.

@@

A second implicit belief regarding the effect of school closure on the spread of COVID-19 is that school not only influences the spread of the virus among children and teenagers but also has a knock-on effect on the spread of the virus in older age groups, also called Secondary Attack Rate (SAR). The contaminated children and teenagers would bring the virus back home and would pass the virus on to their parents and relatives. For example, research investigating the contamination in the household network not only revealed an exceptionally high rate of secondary contamination but also that this contamination happened when the schools were closed [@sorianoarandes2021household]. Despite being reported in several clinical and epidemiological studies [@siebach2021childhood; @zhendong2020clinical], multiple research have shown that the SAR from children to household members was lower than expected [@heavey2020evidence; @vanderhoek2020role; @kim2021role; @ludvigsson2020children]. However, the SAR of children and teenagers to the household member is likely to be age-dependent, with differences between infants, primary and secondary school children, and college students [@grasleguen2021reopening]. If a secondary transmission from children and teenagers to household member has a significant influence, then a temporal causality relationship between their evolution should be observed.

@@

# Methods

## Study design

A longitudinal study was conducted to observe changes in COVID-19 cases during school closures. Changes in COVID-19 from five age groups from the first to the 28th day of school closure are compared across 12 countries. Then, a causality analysis evaluate the impact of changes in these younger groups on older age groups.

## Data collection

Data on COVID-19 case counts by age group have been obtained from the COVerAGE-DB project [@riffe2021data]. The COVerAGE-DB which collect numbers for every countries by groups of 5 or 10 years if available. In addition, spline approximations are used to deals with the heterogeneity of countries’ reporting formats by using when the data for this age bracket is not available for a country.

## Data validation

## Generalized additive model

@@

## Transfer entropy

@@

@@

# Discussion

@@

# Author statements

## Ethical approval

This study is a longitudinal analysis on available data; thus, it does not require ethical approval.

## Funding

None declared.

## Competing interests

None declared.

## Data Availability

The data can be accessed from the COVerAGE-DB OSF repository (https://osf.io/mpwjq/) and from the UNESCO servers (https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip). All preprocessing, analyses and code used to build the submitted paper are available at https://github.com/damien-dupre/covid_ts_causality.

### 2024-02-26 manuscript_elsevier/title_page.Rmd
abstract

---
title: Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
author:
  - name: Damien Dupré
    email: damien.dupre@dcu.ie
    affiliation: Business School, Dublin City University
    correspondingauthor: true
  - name: Edgar Morgenroth
    email: edgar.morgenroth@dcu.ie
    affiliation: Business School, Dublin City University
address:
  - code: Business School, Dublin City University
    organization: Business School, Dublin City University
    city: Dublin
    country: Ireland
abstract: |
  This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups. The study also evaluates the effects of school closures on COVID-19 cases using a Generalized Additive Model (GAM). The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied. However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases. Age groups 15 to 24 show a surge immediately after closure, followed by a decline. Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa. These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
keywords: 
  - COVID-19
  - School Closures
  - Intergenerational Transmission.
journal: "Public Health"
date: "`r Sys.Date()`"
classoption: preprint, 3p
linenumbers: false
numbersections: true
bibliography: bibliography.bib
# biblio-style: elsarticle-num
csl: https://www.zotero.org/styles/elsevier-with-titles
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
    keep_tex: false
    citation_package: default
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{caption}
---

>>>>

---
title: Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
author:
  - name: Damien Dupré
    email: damien.dupre@dcu.ie
    affiliation: Business School, Dublin City University
    correspondingauthor: true
  - name: Edgar Morgenroth
    email: edgar.morgenroth@dcu.ie
    affiliation: Business School, Dublin City University
address:
  - code: Business School, Dublin City University
    organization: Business School, Dublin City University
    city: Dublin
    country: Ireland
abstract: |
  *Objectives:* This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups.

  *Study Design:* A longitudinal study was conducted to observe changes in COVID-19 cases during school closures.

  *Methods:* A Generalized Additive Model is performed to compare trends from twelve European countries across five age groups. Then, Transfer Entropy calculations are used to evaluate the impact of younger cohorts on older cohorts.

  *Results:* The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied. However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases. Age groups 15 to 24 show a surge immediately after closure, followed by a decline. Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa.

  *Conclusions:* These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
keywords: 
  - COVID-19
  - School Closures
  - Intergenerational Transmission.
journal: "Public Health"
date: "`r Sys.Date()`"
classoption: preprint, 3p
linenumbers: false
numbersections: true
bibliography: bibliography.bib
# biblio-style: elsarticle-num
csl: https://www.zotero.org/styles/elsevier-with-titles
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
    keep_tex: false
    citation_package: default
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{caption}
---
